MG7: Configurable and scalable 16S data analysis

### 2016-07-26 docs/mg7-paper/out/paper.tex
note for LCA abbrv

\begin{itemize}
\tightlist
\item
  Static reproducible specification of dependencies and behavior of the
  different components using
  \emph{\href{https://github.com/ohnosequences/statika}{Statika}} and
  \emph{\href{https://github.com/ohnosequences/datasets}{Datasets}}
\item
  Parallelization and distributed analysis based on AWS, with on-demand
  infrastructure as the basic paradigm
\item
  Definition of complex workflows using
  \emph{\href{https://github.com/ohnosequences/loquat}{Loquat}}, a
  composable system for scaling/parallelizing stateless computations
  especially designed for AWS
\item
  A new approach to data analysis specification, management and
  specification based on working with it in exactly the same way as for
  a software project, together with the extensive use of compile-time
  structures and checks
\item
  Modeling of the taxonomy tree using the new paradigm of graph
  databases (Bio4j, \citep{pareja2015bio4j}). It facilitates the
  taxonomic assignment tasks and the calculation of the taxa abundance
  values considering the hierarchic structure of taxonomy tree
\item
  Exhaustive per-read taxonomic assignment using two complementary
  assignment algorithms Lowest Common Ancestor and Best BLAST Hit
\item
  Using a new 16S database of reference sequences (16S-DB7) with a
  flexible and sustainable system of updating and project-driven
  customization
\end{itemize}

@@

\href{https://github.com/ohnosequences/statika}{Statika} is a Scala
library developed by \emph{AA} and \emph{EPT} which serves as a way of
defining and composing machine behaviors statically. The main component
are \textbf{bundles}. Each bundle declares a sequence of computations
(its behavior) which will be executed in an \textbf{environment}. A
bundle can \emph{depend} on other bundles, and when being executed by an
environment, its DAG (Directed Acyclic Graph) of dependencies is
linearized and run in sequence. In our use, bundles correspond to what
an EC2 instance should do and an environment to an AMI (Amazon Machine
Image) which prepares the basic configuration, downloads the Scala code
and runs it.

@@

\href{https://github.com/ohnosequences/datasets}{Datasets} is a Scala
library developed by \emph{AA} and \emph{EPT} with the goal of being a
Scala-embedded mini-language for datasets and their locations.
\textbf{Data} is represented as type-indexed fields: keys are modeled as
singleton types, and values correspond to what could be called a
denotation of the key: a value of type \texttt{Location} tagged with the
key type. Then a \textbf{Dataset} is essentially a collection of data,
which are guaranteed statically to be different through type-level
predicates, making use of the value--type correspondence which can be
established through singleton types and implicits. A dataset location is
then just a list of locations formed by locations of each dataset key.
All this is based on what could be described as an embedding in Scala of
an extensible record system with concatenation on disjoint labels, in
the spirit of \citep{harper1990extensible} \citep{harper1991record}. For
that \emph{Datasets} uses the
\href{https://github.com/ohnosequences/cosas/}{ohnosequences/cosas}
library.

@@

\href{https://github.com/ohnosequences/loquat}{Loquat} is a library
developed by \emph{AA}, \emph{EK} and \emph{EPT} designed for the
execution of embarrassingly parallel tasks using S3, SQS and EC2 Amazon
services.

@@

We developed our own Scala-based type-safe eDSLs (embedded Domain
Specific Languages) for
\href{https://github.com/ohnosequences/flash-api}{FLASh}
\citep{magovc2011flash} and
\href{https://github.com/ohnosequences/blast-api}{BLAST}
\citep{camacho2009blast} expressions and their execution.

@@

The starting point for our 16S reference database is RNA Central
\citep{rnacentral2014rnacentral},
\href{http://blog.rnacentral.org/2016/03/rnacentral-release-5.html}{version
5}. RNACentral was chosen being the most up to date, comprehensive RNA
sequence repository, including among others all RNA data from Silva,
GreenGenes, RDP, ENA, and RefSeq. First we take those sequences which

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  are annotated as being of \texttt{rRNA} type\footnote{We are aware of
    the existence of a gene annotation corresponding to 16S in
    RNACentra, that we are \textbf{not using} due to a significant
    amount of 16S sequences lacking it}
\item
  their length is at least \texttt{1300}\footnote{Note that 16S
    sequences are sometimes part of an entry corresponding to whole
    small subunits; that's why we do not set a maximum length threshold}
\item
  have at least one taxonomic assignment to a descendant of
  \emph{Bacteria} or \emph{Archaea}
\item
  their lineage does not contain a set of taxa deemed
  uninformative\footnote{for example: ``unclassified Bacteria
    (miscellaneous)'' or ``unclassified''. Assigning a read to an
    ``unclassified'' taxon defeats the first and foremost goal of
    taxonomic profiling: \emph{classification}.}
\end{enumerate}

After this first step, we drop redundant assignments: if sequences
\(S_1 \subseteq S_2\) share an assignment \(T\), it gets dropped from
\(S_1\); sequences which as a result end up having no assignments are
removed.

Once we have non-redundant set of RNA sequences containing all those
corresponding to 16S, we can apply our procedure for checking internal
consistency of taxonomic assignments.

We run MG7 using this set of sequences as query, and with reference all
but the one we are trying to assign. Explain

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  which kind of misassignments can be detected
\item
  how
\item
  why it is OK for sequences assigned to strange taxa, with no close 16S
  seq (they will get no hits)
\item
  we analyze every assignment independently
\end{enumerate}

>>>>

\begin{itemize}
\tightlist
\item
  Static reproducible specification of dependencies and behavior of the
  different components using \emph{\protect\hyperlink{statika}{Statika}}
  and \emph{\protect\hyperlink{datasets}{Datasets}}
\item
  Parallelization and distributed analysis based on AWS, with on-demand
  infrastructure as the basic paradigm
\item
  Definition of complex workflows using
  \emph{\protect\hyperlink{loquat}{Loquat}}, a composable system for
  scaling/parallelizing stateless computations especially designed for
  AWS
\item
  A new approach to data analysis specification, management and
  specification based on working with it in exactly the same way as for
  a software project, together with the extensive use of compile-time
  structures and checks
\item
  Modeling of the taxonomy tree using the new paradigm of graph
  databases (Bio4j, \citep{pareja2015bio4j}). It facilitates the
  taxonomic assignment tasks and the calculation of the taxa abundance
  values considering the hierarchic structure of taxonomy tree
\item
  Exhaustive per-read taxonomic assignment using two complementary
  assignment algorithms: Lowest Common Ancestor (LCA) and Best BLAST Hit
\item
  Using a new 16S database of reference sequences (16S-DB7) with a
  flexible and sustainable system of updating and project-driven
  customization
\end{itemize}

@@

\protect\hyperlink{statika}{Statika} is a Scala library developed by
\emph{AA} and \emph{EPT} which serves as a way of defining and composing
machine behaviors statically. The main component are \textbf{bundles}.
Each bundle declares a sequence of computations (its behavior) which
will be executed in an \textbf{environment}. A bundle can \emph{depend}
on other bundles, and when being executed by an environment, its DAG
(Directed Acyclic Graph) of dependencies is linearized and run in
sequence. In our use, bundles correspond to what an EC2 instance should
do and an environment to an AMI (Amazon Machine Image) which prepares
the basic configuration, downloads the Scala code and runs it.

@@

\protect\hyperlink{datasets}{Datasets} is a Scala library developed by
\emph{AA} and \emph{EPT} with the goal of being a Scala-embedded
mini-language for datasets and their locations. \textbf{Data} is
represented as type-indexed fields: keys are modeled as singleton types,
and values correspond to what could be called a denotation of the key: a
value of type \texttt{Location} tagged with the key type. Then a
\textbf{Dataset} is essentially a collection of data, which are
guaranteed statically to be different through type-level predicates,
making use of the value--type correspondence which can be established
through singleton types and implicits. A dataset location is then just a
list of locations formed by locations of each dataset key. All this is
based on what could be described as an embedding in Scala of an
extensible record system with concatenation on disjoint labels, in the
spirit of \citep{harper1990extensible} \citep{harper1991record}. For
that \emph{Datasets} uses the {[}ohnosequences/cosas{]}{[}cosas{]}
library.

@@

\protect\hyperlink{loquat}{Loquat} is a library developed by \emph{AA},
\emph{EK} and \emph{EPT} designed for the execution of embarrassingly
parallel tasks using S3, SQS and EC2 Amazon services.

@@

We developed our own Scala-based type-safe eDSLs (embedded Domain
Specific Languages) for {[}FLASh{]}{[}flash{]} \citep{magovc2011flash}
and {[}BLAST{]}{[}blast{]} \citep{camacho2009blast} expressions and
their execution.

@@

Our 16S-DB7 Reference Database is a curated subset of sequences from the
NCBI nucleotide database \textbf{nt}. The sequences included were
selected by similarity with the bacterial and archaeal reference
sequences downloaded from the \textbf{RDP database}
\citep{cole2013ribosomal}. RDP unaligned sequences were used to capture
new 16S RNA sequences from \textbf{nt} using BLAST similarity search
strategies and then, performing additional curation steps to remove
sequences with poor taxonomic assignments to taxonomic nodes close to
the root of the taxonomy tree. All the nucleotide sequences included in
\textbf{nt} database has a taxonomic assignment provided by the
\textbf{Genbank} sequence submitter. NCBI provides a table (available at
ftp://ftp.ncbi.nlm.nih.gov/pub/taxonomy/) to do the mapping of any
Genbank Identifier (GI) to its Taxonomy Identifier (TaxID). Thus, we are
based on a crowdsourced submitter-maintained taxonomic annotation system
for reference sequences. It supposes a sustainable system able to face
the expected number of reference sequences that will populate the public
global nucleotide databases in the near future. Another advantageous
point is that we are based on NCBI taxonomy, the \emph{de-facto}
standard taxonomic classification for biomolecular data
\citep{cochrane20102010}. NCBI taxonomy is, undoubtedly, the most used
taxonomy all over the world and the most similar to the official
taxonomies of each specific field. This is a crucial point because all
the type-culture and tissue databanks follow this official taxonomical
classification and, in addition, all the knowledge accumulated during
last decades is referred to this taxonomy. In addition NCBI provides a
direct connection between taxonomical formal names and the physical
specimens that serve as exemplars for the species
\citep{federhen2014type}.

Certainly, if metagenomics results are easily integrated with the
theoretical and experimental knowledge of each specific area, the impact
of metagenomics will be higher than if it progresses as a disconnected
research branch. Considering that metagenomics data interoperability,
which is especially critical in clinical environments, requires a stable
taxonomy to be used as reference, we decided to rely on the most widely
used taxonomy: the NCBI taxonomy. In addition, the biggest global
sequence database GenBank follows this taxonomy to register the origin
of all their submitted sequences. Our 16S database building strategy
allows the substitution of the 16S database by any other subset of
\textbf{nt}, even by the complete \textbf{nt} database if it would be
needed, for example, for analyzing shotgun metagenomics data. This
possibility of changing the reference database provides flexibility to
the system enabling it for easy updating and project-driven
personalization.
